Sensor Clusterization in D-optimal Design in Infinite Dimensional Linear Bayesian Inverse Problems

### 2020-09-30 new_paper.tex
backup, trying to diff

\title[Sensor Clusterization in D-optimal design in infinite
  dimensions]{Sensor Clusterization in D-optimal Design in Infinite
  Dimensional \reftwo{Linear} Bayesian Inverse Problems}

@@

\begin{abstract}
  We investigate the problem of sensor clusterization in optimal
  experimental design for infinite-dimensional Bayesian
  \reftwo{linear} inverse problems. We suggest an analytically
  tractable model for such designs and reason how it may lead to
  sensor clusterization in the case of iid measurement noise. We also
  show that in the case of spatially correlated measurement error
  clusterization does not occur. As a part of the analysis we prove a
  matrix determinant lemma analog in infinite dimensions, as well as a
  lemma for calculating derivatives of $\log \det$ of operators.
\end{abstract}

@@

Designing experiments is usually done by optimizing some \emph{design
  criterion} and this is true both for frequentists
\cite{Silvey13,Ucinski05} as well as for Bayesians
\cite{ChalonerVerdinelli95}, see \cite{ChalonerVerdinelli95} for an
investigation of the analogy between the two approaches. Although
there is a plethora of design criteria, in the current work we will
focus on just one of these, commonly referred to as \emph{D-optimal
  design}. It has a simple and appealing motivation in the Bayesian
context as explained in \cite{ChalonerVerdinelli95}: for a linear
model, a D-optimal design seeks to maximize the expected information
gain (KL divergence \cite{KullbackLeibler51,CoverThomas12}) between
posterior and prior. In finite dimensions, maximizing the expected
information gain amounts to minimizing the determinant of the
posterior covariance matrix. In a frequentist setting\reftwo{\sout{s}}, an optimal
design minimizes the volume of the uncertainty ellipsoid \cite[page
  16]{Ucinski05}, but this is done for the Fisher information matrix
and not the posterior covariance. However, \cite{ChalonerVerdinelli95}
shows that the latter is just a regularized version of the former.

@@

In the process\reftwo{,} we generalize several lemmas from linear algebra to
infinite-dimensional settings --- the Matrix Determinant Lemma and a
lemma for calculating the derivative of determinants. We also provide
other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemmas \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}. %%  We also show that in our model with independent measurement
%% noise, an interesting phenomenon arises, in which the eigenvectors
%% of the measurement operator (denoted below $\obs^*\obs$) are the
%% same as the eigenvectors of the prior covariance in observation
%% space, which helps arrive at the results outlined in the previous
%% paragraph.
The arguments presented are completely generic and apply equally to
any of the scenarios mentioned in the introduction.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Shown are pointwise posterior standard deviations for
    various numbers of sensors, and location of these sensors for the
    1D heat equation with \reftwo{a} homogeneous Dirichlet boundary described in
    section \ref{subsec:example}. Note the clusterization that occurs
    when 6 sensors are used --- only 4 measurement locations are
    included. Observed data does not influence the pointwise
    variances, hence it is omitted.}
\end{figure}\label{fig:clusterization illustration}
In section \ref{section:abstract OED} we present a more abstract and
general formulation of the problem, but for the purpose of
illustration, we present the problem via a toy model --- the 1D heat
equation in $[0,\pi]$ with \reftwo{a} homogeneous Dirichlet boundary
conditions.

@@

We would like to infer the initial condition $u_0$. For that purpose,
we measure $u$ at some set of locations $\x_j \in [0,\pi], j=1, \dots,m$
and a final time $T > 0$. We assume centered Gaussian measurement
error, so we can observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2), \sigma > 0$ iid. We model the
initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with \reftwo{a} homogeneous Dirichlet boundary condition. It is
well known \cite{Tarantola05} that for linear problems, with Gaussian
prior and error, the posterior is also Gaussian with a covariance that
does not depend on the observed data. The posterior covariance
$\postcov$ is known to have a closed form formula, even in infinite
dimensions\cite{Stuart10}. If we denote by $\fwd$ the dynamics
operator, so that $u( \cdot,T) = \fwd u_0$, and the observation
operator $\obs$ so that $u(\x_j,T) = (\obs u)_j, j=1,\dots,m$, then the
posterior covariance is known and depends only on $\prcov, \fwd, \obs$ and
$\sigma^2$ (see section \ref{section:abstract OED} and
\eqref{eq:postcov} specifically).

@@

\section{Preliminaries and Notation}\label{section:prelim}
\reftwo{
\subsection{Overview}
The theoretical foundations for inverse problems over function spaces
can be found in \cite{Stuart10}. In this section we define notations that will be
  used throughout the paper.
%% A reader mostly familiar with the
%% literature may initially want to skip this section and only consult
%% the list below, which summarizes the notation used. We denote:
%% \begin{itemize}
%%   \item Parameter space $\hil$.
%%   \item Data space $\ban$.
%%   \item Forward dynamics operator $\fwd: \hil \to \ban$.
%%   \item Observation operator $\obs: \ban \to \mathbb{R}^m$.
%%   \item Prior measure $\pr = \mathcal{N}(0, \prcov)$.
%%   \item Data $\data \in \mathbb{R}^m$.
%%   \item Observation and model errors, captured in $\Sigma(\obs)$.
%% \end{itemize}
%% The design objective was derived in \cite{AlexanderianGloorGhattas14}
%% and recorded in \eqref{eq:objective} below.
\sout{At this point, it is useful to record a known result regarding
  the posterior covariance operator of our inverse problem, namely}
\begin{align}\label{eq:postcov precursor}
  \stkout{\postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd )^{-1}.}
\end{align}
}

\subsection{Bayesian \reftwo{Linear} Inverse Problems}\label{section:abstract OED}
Let $\hil$ be a Hilbert space and $\ban$ a Banach space, both
infinite-dimensional. Consider $\fwd: \hil \to \ban$, a linear
operator (the ``forward operator''). We are interested in forward
operators that are strongly smoothing (have fast decaying singular
values --- the heat operator from section \ref{subsec:example} is a
prime example). We want to infer $\param \in \hil$, some parameter of
the dynamics given noisy observations of $\fwd \param$. \reftwo{We
  take a zero mean Gaussian prior $\pr$ for $\param$ (otherwise the
  mean can just be subtracted everywhere). Hence, \sout{We model}}
$\param \sim \pr = \normal(0,\prcov)$ with some appropriate covariance
operator $\prcov$ \cite{Stuart10}. Observations (measurements) are
taken via the linear observation operator $\obs \in ( \ban^* )^m$,
where $m$ is the number of observations allowed (e.g. sensors at our
disposal). In an analogy with linear algebra (where row vectors are
thought of as linear functionals), we think of $\obs$ as having
``rows'' $\meas_j, j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \ban^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \ban$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
A few observations regarding $\obs$ are in order. First, it is good to
keep in mind that $( \ban^* )^m$ is a Banach space with norm $\| \obs
\| = \sum_{j=1}^m \|\meas_j\|$. Next, for $u \in \ban$ and $v\in
\R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

@@

\subsubsection{Error Terms}
Considering the sum of the error terms\reftwo{,} it is easy to see that $\obs
\eps' + \eps \in \R^m$. It is a centered Gaussian random vector and
its covariance matrix is
\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps)  (\obs \eps' + \eps)^t ] 
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I , 
  \end{split}
\end{align}
where
\begin{equation}\label{eq:modcov explained}
  [\obs \modcov \obs^*]_{ij} = e_i^t \obs \modcov \obs^* e_j = \meas_i (\modcov \meas_j),
\end{equation}
where the first equality in \eqref{eq:modcov explained} holds by
definition and the second by \eqref{eq:obs*}. The explicit dependence
on $\obs$ will be mostly dropped for notational convenience, so
$\Sigma(\obs) = \Sigma$. Thus, for a fixed $\obs$ (i.e.\ a fixed set
of measurements) we can equivalently write $\data = \obs \fwd \param +
\bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$. At this point,
it is useful to \reftwo{\sout{rewrite the posterior covariance operator of our
inverse problem from \eqref{eq:postcov precursor}} record a known result
regarding the posterior covariance operator of our inverse problem,
namely}
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}
\refthree{
Usually, this is written using $\tmp := \obs \fwd$ and $\tmp^*$ but
since we need to separate the observation operator $\obs$ from the
forward operator $\fwd$ we write the explicit expression
\eqref{eq:postcov}.}

@@

\subsection{D-Optimal Design in Infinite Dimensions}\label{subsec:D optimal design} 
The meaning of D-optimal design in infinite-dimensional Hilbert spaces
was investigated in \cite{AlexanderianGloorGhattas14}. Note that the
authors make assumptions that amount to $\Sigma=I$ (implied by
$\modcov = 0,\sigma^2=1$), but we choose not to take these
simplification here. This is because $\Sigma$ can determine ``how
much'' clusterization we see. As stated
\cite[pp. 681]{AlexanderianGloorGhattas14}, the results hold for more
general covariance matrices. The conclusion is that in
infinite-dimensions, a D-optimal design is well-defined as maximizing
the expected KL divergence between posterior and prior
\reftwo{(denoted $\tar$ below and throughout the paper)} and is summarized, using our notation,
in the following theorem:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]
  Let $\pr = \normal(0,\prcov)$ be a centered Gaussian prior
  on $\hil$   and let $\post = \normal(\postmean,\postcov)$ 
  the posterior measure on $\ban$ for the Bayesian linear
  inverse problem $\data =  \obs \fwd\param + \obs \eps' + \eps$ discussed
  above. Then 
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

\subsection{\reftwo{Overview}}
From \eqref{eq:objective} we wish to characterize solution(s) of the
following optimization problem for $\tar$. %%: (\ban^*)^m \to \R$:
\begin{align}\label{eq:optimization}
  \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
  = \argmax_{\obs} \frac12 \log \det 
  (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
\end{align}
where $\obs$ is constrained to some allowed set of measurements
(e.g.\ pointwise evaluations on $[0,\pi]$).
%%, for the example in section \ref{subsec:example}).
We seek a formulation of the optimal design using Lagrange
multipliers. In section \ref{section:objective} we find the gradient
of $\tar$ and record it in \eqref{eq:tar grad}. In section
\ref{subsec:unit length} we relax the constraints on $\obs$ so that
the optimization problem in \eqref{eq:optimization} \reftwo{is
  analytically tractable}.  We find gradients for the new constraints
\reftwo{and record them in \eqref{eq:grad constraints}. Finally, in
  section \ref{subsec:necessary} we use these to formulate the optimal
  design problem as a Lagrange multipliers problem in equation
  \eqref{eq:conditions}.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis of Optimal Designs --- Vanishing Model Error}\label{section:vanishing}
\subsection{\reftwo{Overview}}
\reftwo{In this section we analyze optimal designs when no model error
  is present. We show how sensor clusterization arises in section
  \ref{subsec:clusterization}. We prepare the ground by first deriving
  an eigenvalue problem in $\obs$ for an optimal design in section
  \ref{subsec:eigenproblem}. This eigenvalue problem is recorded in
  \eqref{eq:mod conditions}. Next, we characterize optimal designs as
  ones that make all eigenvalues in equal in the above mentioned
  eigenproblem. This is crucial for clusterization to occur. The
  derivation is presented in section \ref{subsec:characterization}. }

\subsection{\reftwo{The Eigenvalue Problem}}\label{subsec:eigenproblem}
When $\modcov = 0$, then $\Sigma= \sigma^2I$. The necessary
first-order condition for optimality \eqref{eq:conditions} becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi, 
\end{equation}
with $\Xi$ diagonal. This is an eigenvalue problem for the
self-adjoint operator $\sigma^{-2}\fwd \postcov \fwd^*$ where the rows
of $\obs$, namely $\meas_j,j=1,\dots, m$, are the
eigenvectors. \refthree{Below, we use \eqref{eq:postcov} in the second
  equality.} If $\fwd$ is invertible, \refthree{\sout{then we may
    write} the third equality is trivial. Otherwise, we use some
  calculations delegated to lemma \ref{lemma:twice woodbury} in the
  appendix}:
\begin{align}\label{eq:mod conditions}
  \begin{split}
  \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \\
  %
  %
  %
  &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^* \\
  %
  %
  %
  &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^*
  \end{split}
\end{align}  
\refthree{\sout{which} This} presents the optimal design problem as a
problem on $\ban$ and not on $\hil$ --- the term $(\fwd \prcov
\fwd^*)^{-1}$ is the prior precision in $\ban$ and \reftwo{\sout{the
    big parentheses enclose} $(\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}
  \obs^* \obs$ is} the posterior precision in
$\ban$. \refthree{\sout{This expression makes sense even if $\fwd$ is
    not invertible --- consider instead $\fwd$ restricted to $\ker
    \fwd^{\perp}$. This does not change the optimal design problem at
    all, since we will never take measurements that have a component
    in $\ker \fwd$ (this can be understood from Corollary
    \ref{cor:zero mod err}).}}

@@

Thus, if we take $\func_j^{*} = \meas_j$ and $C := (\fwd \prcov
\fwd^*)^{-1}$ in the above lemma, we conclude from the eigenproblem
\eqref{eq:mod conditions} that $\obs^*\obs$ has the same eigenvectors
as $\fwd \prcov \fwd^*$, namely $\{\ev_i\}_{i=1}^{\infty}$ (infinitely
many of them with zero eigenvalue, since $\obs^*\obs$ is
finite-rank). Denote the corresponding non-zero eigenvalues of
$\obs^*\obs$ by $\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i
\geq k+1$. \refthree{\stkout{By the argument in the paragraph following
    \eqref{eq:mod conditions}:}}

\refthree{
  \begin{align}%\label{eq:old pre manipulations}
    \cancel{
      \begin{split}
        I + \sigma^{-2} \prcov^{1/2} \fwd ^* \obs^*
        \obs \fwd \prcov^{1/2} &= \prcov^{1/2} ( \prcov^{-1} +
        \sigma^{-2} \fwd^*\obs^* \obs \fwd ) \prcov^{1/2} \\
        %
        %
        %
        &=\prcov^{1/2} \fwd^* ( \fwd^{-*}
        \prcov^{-1}\fwd^{-1} + \sigma^{-2} \obs^* \obs ) \fwd \prcov^{1/2}
        (\fwd\prcov\fwd^*)^{-1} \\
        %
        %
        %
        &= \prcov^{1/2} \fwd^* \left ( (\fwd \prcov \fwd^* )^{-1}
        +\sigma^{-2} \obs^* \obs \right ) \fwd \prcov^{1/2}.
      \end{split}
    }
\end{align}
}

\refthree{
\begin{align}\label{eq:pre manipulations}
  \begin{split}
    \det \left (I + \sigma^{-2} \prcov^{1/2} \fwd ^* \obs^*
    \obs \fwd \prcov^{1/2}\right ) &=
    %
    \det \left (I + \sigma^{-2} \obs^*
    \obs \fwd \prcov\fwd^* \right ) \text{(Sylvester's determinant theorem)}\\
    %
    %
    %
    &= \det \left \{ \left ( (\fwd \prcov \fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs \right ) (\fwd \prcov \fwd^*)\right \}\\
  \end{split}
\end{align}
}

@@

\reftwo{Results of section \ref{subsec:characterization} specifically,
  the discussion at the end of said section, along with figure
  \ref{figure:clusterization} and equation \eqref{eq:true target} show
  how clusterization arises when an optimal design is saught.

>>>>

\title[Sensor Clusterization in D-optimal design in infinite
  dimensions]{Sensor Clusterization in D-optimal Design in Infinite
  Dimensional Linear Bayesian Inverse Problems}

@@

\begin{abstract}
  We investigate the problem of sensor clusterization in optimal
  experimental design for infinite-dimensional Bayesian
  linear inverse problems. We suggest an analytically
  tractable model for such designs and reason how it may lead to
  sensor clusterization in the case of iid measurement noise. We also
  show that in the case of spatially correlated measurement error
  clusterization does not occur. As a part of the analysis we prove a
  matrix determinant lemma analog in infinite dimensions, as well as a
  lemma for calculating derivatives of $\log \det$ of operators.
\end{abstract}

@@

Designing experiments is usually done by optimizing some \emph{design
  criterion} and this is true both for frequentists
\cite{Silvey13,Ucinski05} as well as for Bayesians
\cite{ChalonerVerdinelli95}, see \cite{ChalonerVerdinelli95} for an
investigation of the analogy between the two approaches. Although
there is a plethora of design criteria, in the current work we will
focus on just one of these, commonly referred to as \emph{D-optimal
  design}. It has a simple and appealing motivation in the Bayesian
context as explained in \cite{ChalonerVerdinelli95}: for a linear
model, a D-optimal design seeks to maximize the expected information
gain (KL divergence \cite{KullbackLeibler51,CoverThomas12}) between
posterior and prior. In finite dimensions, maximizing the expected
information gain amounts to minimizing the determinant of the
posterior covariance matrix. In a frequentist setting, an optimal
design minimizes the volume of the uncertainty ellipsoid \cite[page
  16]{Ucinski05}, but this is done for the Fisher information matrix
and not the posterior covariance. However, \cite{ChalonerVerdinelli95}
shows that the latter is just a regularized version of the former.

@@

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings --- the Matrix Determinant Lemma and a
lemma for calculating the derivative of determinants. We also provide
other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemmas \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}. %%  We also show that in our model with independent measurement
%% noise, an interesting phenomenon arises, in which the eigenvectors
%% of the measurement operator (denoted below $\obs^*\obs$) are the
%% same as the eigenvectors of the prior covariance in observation
%% space, which helps arrive at the results outlined in the previous
%% paragraph.
The arguments presented are completely generic and apply equally to
any of the scenarios mentioned in the introduction.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Shown are pointwise posterior standard deviations for
    various numbers of sensors, and location of these sensors for the
    1D heat equation with homogeneous Dirichlet boundary described in
    section \ref{subsec:example}. Note the clusterization that occurs
    when 6 sensors are used --- only 4 measurement locations are
    included. Observed data does not influence the pointwise
    variances, hence it is omitted.}
\end{figure}\label{fig:clusterization illustration}
In section \ref{section:abstract OED} we present a more abstract and
general formulation of the problem, but for the purpose of
illustration, we present the problem via a toy model --- the 1D heat
equation in $[0,\pi]$ with homogeneous Dirichlet boundary
conditions.

@@

We would like to infer the initial condition $u_0$. For that purpose,
we measure $u$ at some set of locations $\x_j \in [0,\pi], j=1, \dots,m$
and a final time $T > 0$. We assume centered Gaussian measurement
error, so we can observe $v(\x_j,T) = u(\x_j,T) + \eps(\x_j)$ with
$\eps(\x_j) \sim \normal(0, \sigma^2), \sigma > 0$ iid. We model the
initial condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with homogeneous Dirichlet boundary condition. It is
well known \cite{Tarantola05} that for linear problems, with Gaussian
prior and error, the posterior is also Gaussian with a covariance that
does not depend on the observed data. The posterior covariance
$\postcov$ is known to have a closed form formula, even in infinite
dimensions\cite{Stuart10}. If we denote by $\fwd$ the dynamics
operator, so that $u( \cdot,T) = \fwd u_0$, and the observation
operator $\obs$ so that $u(\x_j,T) = (\obs u)_j, j=1,\dots,m$, then the
posterior covariance is known and depends only on $\prcov, \fwd, \obs$ and
$\sigma^2$ (see section \ref{section:abstract OED} and
\eqref{eq:postcov} specifically).

@@

\section{Preliminaries and Notation}\label{section:prelim}

\subsection{Overview}
The theoretical foundations for inverse problems over function spaces
can be found in \cite{Stuart10}. In this section we define notations that will be
  used throughout the paper.
%% A reader mostly familiar with the
%% literature may initially want to skip this section and only consult
%% the list below, which summarizes the notation used. We denote:
%% \begin{itemize}
%%   \item Parameter space $\hil$.
%%   \item Data space $\ban$.
%%   \item Forward dynamics operator $\fwd: \hil \to \ban$.
%%   \item Observation operator $\obs: \ban \to \mathbb{R}^m$.
%%   \item Prior measure $\pr = \mathcal{N}(0, \prcov)$.
%%   \item Data $\data \in \mathbb{R}^m$.
%%   \item Observation and model errors, captured in $\Sigma(\obs)$.
%% \end{itemize}
%% The design objective was derived in \cite{AlexanderianGloorGhattas14}
%% and recorded in \eqref{eq:objective} below.

\subsection{Bayesian Linear Inverse Problems}\label{section:abstract OED}
Let $\hil$ be a Hilbert space and $\ban$ a Banach space, both
infinite-dimensional. Consider $\fwd: \hil \to \ban$, a linear
operator (the ``forward operator''). We are interested in forward
operators that are strongly smoothing (have fast decaying singular
values --- the heat operator from section \ref{subsec:example} is a
prime example). We want to infer $\param \in \hil$, some parameter of
the dynamics given noisy observations of $\fwd \param$. We take a zero
mean Gaussian prior $\pr$ for $\param$ (otherwise the mean can just be
subtracted everywhere). Hence, $\param \sim \pr = \normal(0,\prcov)$
with some appropriate covariance operator $\prcov$
\cite{Stuart10}. Observations (measurements) are taken via the linear
observation operator $\obs \in ( \ban^* )^m$, where $m$ is the number
of observations allowed (e.g. sensors at our disposal). In an analogy
with linear algebra (where row vectors are thought of as linear
functionals), we think of $\obs$ as having ``rows'' $\meas_j,
j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \ban^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \ban$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
A few observations regarding $\obs$ are in order. First, it is good to
keep in mind that $( \ban^* )^m$ is a Banach space with norm $\| \obs
\| = \sum_{j=1}^m \|\meas_j\|$. Next, for $u \in \ban$ and $v\in
\R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

@@

\subsubsection{Error Terms}
Considering the sum of the error terms it is easy to see that $\obs
\eps' + \eps \in \R^m$. It is a centered Gaussian random vector and
its covariance matrix is
\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps)  (\obs \eps' + \eps)^t ] 
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I , 
  \end{split}
\end{align}
where
\begin{equation}\label{eq:modcov explained}
  [\obs \modcov \obs^*]_{ij} = e_i^t \obs \modcov \obs^* e_j = \meas_i (\modcov \meas_j),
\end{equation}
where the first equality in \eqref{eq:modcov explained} holds by
definition and the second by \eqref{eq:obs*}. The explicit dependence
on $\obs$ will be mostly dropped for notational convenience, so
$\Sigma(\obs) = \Sigma$. Thus, for a fixed $\obs$ (i.e.\ a fixed set
of measurements) we can equivalently write $\data = \obs \fwd \param +
\bar{\eps}$ with $\bar{\eps} \sim \normal(0,\Sigma)$. At this point,
it is useful to record a known result regarding the posterior
covariance operator of our inverse problem, namely
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}
Usually, this is written using $\tmp := \obs \fwd$ and $\tmp^*$ but
since we need to separate the observation operator $\obs$ from the
forward operator $\fwd$ we write the explicit expression
\eqref{eq:postcov}.

@@

\subsection{D-Optimal Design in Infinite Dimensions}\label{subsec:D optimal design} 
The meaning of D-optimal design in infinite-dimensional Hilbert spaces
was investigated in \cite{AlexanderianGloorGhattas14}. Note that the
authors make assumptions that amount to $\Sigma=I$ (implied by
$\modcov = 0,\sigma^2=1$), but we choose not to take these
simplification here. This is because $\Sigma$ can determine ``how
much'' clusterization we see. As stated
\cite[pp. 681]{AlexanderianGloorGhattas14}, the results hold for more
general covariance matrices. The conclusion is that in
infinite-dimensions, a D-optimal design is well-defined as maximizing
the expected KL divergence between posterior and prior (denoted $\tar$
below and throughout the paper) and is summarized, using our notation,
in the following theorem:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]
  Let $\pr = \normal(0,\prcov)$ be a centered Gaussian prior
  on $\hil$   and let $\post = \normal(\postmean,\postcov)$ 
  the posterior measure on $\ban$ for the Bayesian linear
  inverse problem $\data =  \obs \fwd\param + \obs \eps' + \eps$ discussed
  above. Then 
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

\subsection{Overview}
From \eqref{eq:objective} we wish to characterize solution(s) of the
following optimization problem for $\tar$. %%: (\ban^*)^m \to \R$:
\begin{align}\label{eq:optimization}
  \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
  = \argmax_{\obs} \frac12 \log \det 
  (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
\end{align}
where $\obs$ is constrained to some allowed set of measurements
(e.g.\ pointwise evaluations on $[0,\pi]$).
%%, for the example in section \ref{subsec:example}).
We seek a formulation of the optimal design using Lagrange
multipliers. In section \ref{section:objective} we find the gradient
of $\tar$ and record it in \eqref{eq:tar grad}. In section
\ref{subsec:unit length} we relax the constraints on $\obs$ so that
the optimization problem in \eqref{eq:optimization} is analytically
tractable.  We find gradients for the new constraints and record them
in \eqref{eq:grad constraints}. Finally, in section
\ref{subsec:necessary} we use these to formulate the optimal design
problem as a Lagrange multipliers problem in equation
\eqref{eq:conditions}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis of Optimal Designs --- Vanishing Model Error}\label{section:vanishing}
\subsection{Overview}
In this section we analyze optimal designs when no model error is
present. We show how sensor clusterization arises in section
\ref{subsec:clusterization}. We prepare the ground by first deriving
an eigenvalue problem in $\obs$ for an optimal design in section
\ref{subsec:eigenproblem}. This eigenvalue problem is recorded in
\eqref{eq:mod conditions}. Next, we characterize optimal designs as
ones that make all eigenvalues in equal in the above mentioned
eigenproblem. This is crucial for clusterization to occur. The
derivation is presented in section \ref{subsec:characterization}.

\subsection{The Eigenvalue Problem}\label{subsec:eigenproblem}
When $\modcov = 0$, then $\Sigma= \sigma^2I$. The necessary
first-order condition for optimality \eqref{eq:conditions} becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi, 
\end{equation}
with $\Xi$ diagonal. This is an eigenvalue problem for the
self-adjoint operator $\sigma^{-2}\fwd \postcov \fwd^*$ where the rows
of $\obs$, namely $\meas_j,j=1,\dots, m$, are the eigenvectors. Below,
we use \eqref{eq:postcov} in the second equality. If $\fwd$ is
invertible, the third equality is trivial. Otherwise, we use some
calculations delegated to lemma \ref{lemma:twice woodbury} in the
appendix:
\begin{align}\label{eq:mod conditions}
  \begin{split}
  \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \\
  %
  %
  %
  &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^* \\
  %
  %
  %
  &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^*
  \end{split}
\end{align}  
This presents the optimal design problem as a problem on $\ban$ and
not on $\hil$ --- the term $(\fwd \prcov \fwd^*)^{-1}$ is the prior
precision in $\ban$ and $(\fwd\prcov\fwd^*)^{-1} + \sigma^{-2} \obs^*
\obs$ is the posterior precision in $\ban$.

@@

Thus, if we take $\func_j^{*} = \meas_j$ and $C := (\fwd \prcov
\fwd^*)^{-1}$ in the above lemma, we conclude from the eigenproblem
\eqref{eq:mod conditions} that $\obs^*\obs$ has the same eigenvectors
as $\fwd \prcov \fwd^*$, namely $\{\ev_i\}_{i=1}^{\infty}$ (infinitely
many of them with zero eigenvalue, since $\obs^*\obs$ is
finite-rank). Denote the corresponding non-zero eigenvalues of
$\obs^*\obs$ by $\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i
\geq k+1$.

\begin{align}\label{eq:pre manipulations}
  \begin{split}
    \det \left (I + \sigma^{-2} \prcov^{1/2} \fwd ^* \obs^*
    \obs \fwd \prcov^{1/2}\right ) &=
    %
    \det \left (I + \sigma^{-2} \obs^*
    \obs \fwd \prcov\fwd^* \right ) \text{(Sylvester's determinant theorem)}\\
    %
    %
    %
    &= \det \left \{ \left ( (\fwd \prcov \fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs \right ) (\fwd \prcov \fwd^*)\right \}\\
  \end{split}
\end{align}

@@

Results of section \ref{subsec:characterization} specifically, the
discussion at the end of said section, along with figure
\ref{figure:clusterization} and equation \eqref{eq:true target} show
how clusterization arises when an optimal design is saught.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Non-Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis of Optimal Designs --- Non-Vanishing Model Error}
\subsection{Overview}
In this section we show how clusterization is avoided when model
error does not vanish.
